---
title: "Linear modeling of expression in IPF core samples"
author: "[Alex Abbas](mailto:abbas@gene.com)"
date: "`r Sys.Date()`"
output:
    html_document:
      toc: yes
      toc_depth: 4
      theme: flatly
      css : http://research.gene.com/people/abbas/css/tocWide.css
---

```{r global_options, include=F, echo=F}
knitr::opts_chunk$set(fig.width=6, fig.height=6, fig.path='Figs/', echo=F, warning=FALSE, message=FALSE, include=T)
```

```{r load_packages, include=F, echo=F}
library(knitr)
library(Biobase)
library(DT)
library(ggplot2)
library(reshape)
#library(GGally)
#library(ellipse)
#library(RColorBrewer)
```

```{r package_options, include=F, echo=F}
knitr::opts_knit$set(progress = TRUE, verbose = TRUE)
# library(rmarkdown)
# library(BiocStyle)
# render('linearModel.Rmd', "html_document", '../results/linearModel.html')
```

```{r etc, include=F, echo=F}
calcSignatureScores = function(dat=v,geneids=rownames(dat)) {
  if (length(geneids)==1) {
   return(dat[geneids,])
  }
  sliceids = geneids[geneids %in% rownames(dat)]
  sigdat = dat[sliceids,]
  pc1 = prcomp(t(sigdat))$x[,1]
  meansig = apply(sigdat,2,mean)
  if ( cor(pc1,meansig) < 0 ) { pc1 = - pc1 }
  pc1 = pc1 * sd(meansig)/sd(pc1)
  pc1 = pc1 + mean(meansig) - mean(pc1)
  names(pc1) = colnames(dat)
  pc1
}
myminor = unlist(lapply(c(-3:5),function(o){c(2:9)*10^o}))
cv = function(x) { sd(x)/mean(x) }
zscore = function(x) { (x - mean(x,na.rm=T)) / sd(x,na.rm=T) }
```

Hypothesis
===
Histopathological variables are related to differential expression. Build linear models to find these relationships. All analysis on this page is of log2-transformed RPKM data of *IPF patients only*.

Analysis
========
```{r loadData, echo=F, include=F}
load("../data/eset.RData")
pheno = pData(eset)
ipf = eset[,pheno$TISSUE_DIAGNOSIS=="IPF"]
pheno = pData(ipf)
exprs(ipf) = log2(exprs(ipf))

histVars = c(
 "AlveolarSurfaceDensityperMicro",
 "Percent_VolumeFraction_TotalCollagen",
 "AshcroftFibrosisScore_histology",
 "No_TB_permL_MicroCT",
 "Tissue_MicroCT",
 "Neutrophil",
 "Macrophage_CD68",
 "Bcell_CD79a",
 "CD4",
 "CD8",
 "Elastin",
 "Col1",
 "VvCol3",
 "Eosinophils"
)
zHist = do.call(cbind,lapply(histVars,function(histVar){ zscore(pheno[,histVar]) } ))
colnames(zHist) = paste0(histVars,"_z")
rownames(zHist) = rownames(pheno)
phenoz = cbind(pheno,zHist)
pData(ipf) = phenoz
geneids = featureNames(ipf)[fData(ipf)$type=="protein_coding"]
ipf = ipf[geneids,]
```

Examine distributions of vars
-------------
I log-transformed and z-scored all continuous phenotypic variables to prepare for linear modeling. Let's verify that the distributions now look normal.

```{r normal1, fig.width=9, fig.height=7, echo=F}
zHistMelt = melt(phenoz,id.vars=1,measure.vars=colnames(zHist)[1:7])
ggplot(zHistMelt, aes(value, color=variable)) + geom_density()
```
```{r normal2, fig.width=9, fig.height=7, echo=F}
zHistMelt = melt(phenoz,id.vars=1,measure.vars=colnames(zHist)[8:14])
ggplot(zHistMelt, aes(value, color=variable)) + geom_density()
```


Model expression by cell type
-------------

We have counts of six leukocyte types, let's model expression by the z-scores of those. Criteria: p < 0.05, absolute value of logFC > 0.5. Here log fold change is weird, it's actually the log2 expression change per standard deviation unit of the histology variable.

Neutrophils, CD4 T cells, and eosinophils showed no differential expression.

```{r leukfit, echo=F}
pCutoff = 0.05
lfcCutoff = 0.5
design = model.matrix(~Neutrophil_z+Macrophage_CD68_z+Bcell_CD79a_z+CD4_z+CD8_z+Eosinophils_z, ipf) # create a design for linear modelling
fit = eBayes(lmFit(ipf, design)) # run the model
leuks = c("Neutrophil_z","Macrophage_CD68_z","Bcell_CD79a_z","CD4_z","CD8_z","Eosinophils_z")
leuknames = c("Neut","Mac","Bcell","CD4","CD8","Eos")
tts = lapply(leuks,function(leuk){ topTable(fit, coef=leuk, n=100000, p.value=1, lfc=0)[geneids,] })
names(tts) = leuknames
```

```{r updown, echo=F, fig.width=8}
updown = do.call(cbind,lapply(tts, function(sf) {
 oneUpdown = rep(NA,times=dim(sf)[1])
 oneUpdown[ sf$adj.P.Val < pCutoff & sf$logFC > lfcCutoff ] = "up"
 oneUpdown[ sf$adj.P.Val < pCutoff & sf$logFC < -(lfcCutoff)  ] = "down"
 names(oneUpdown) = rownames(sf)
 oneUpdown
}))
colnames(updown) = leuknames
updownTable = do.call(cbind,apply(updown,2,table))
updownTable[2,] = -updownTable[2,]
mycol = c("lightpink","lightblue")
midpoints = barplot(updownTable, beside=TRUE, ylab="number of genes up (down)",col=mycol)
adjustLabelY = function(vec) {
 sapply(vec,function(x){
  delta = 20
  if (x < 0) { delta = -delta }
  if (abs(x) < 50) { delta = -delta }
  x - delta 
 })
}
text(midpoints, adjustLabelY(updownTable), labels=abs(updownTable),cex=.8)
abline(h=0,col="gray")
```

```{r topgenes, echo=F}
tops = lapply(tts,function(tt){
 ttsig = subset(tt,adj.P.Val < pCutoff & abs(logFC)>lfcCutoff)
 tttop = ttsig[order(ttsig$logFC,decreasing=T),]
 data.frame(
  ID = rownames(tttop),
  tttop[,c("symbol","desc","logFC","adj.P.Val")]
 )
})
```

Macrophages
-------------

```{r macs, echo=F}
datatable(tops[[2]]) %>%
  formatRound('logFC', 3) %>%
  formatRound('adj.P.Val', 3)
```

B cells
-------------

```{r b, echo=F}
datatable(tops[[3]]) %>%
  formatRound('logFC', 3) %>%
  formatRound('adj.P.Val', 3)
```

CD8 T cells
-------------

```{r cd8, echo=F}
datatable(tops[[5]]) %>%
  formatRound('logFC', 3) %>%
  formatRound('adj.P.Val', 3)
```

Modules & Info
==============
```{r furlaniModules}
system("module list 2>&1",intern=T)
```

```{r sessioninfo}
sessionInfo()
```




